---
title: "Predictive Modeling"
format: 
  html:
    embed-resources: true
execute:
  warning: false
  message: false
---

## 1. Setup and Data Loading

```{r setup}
#| cache: false
library(tidymodels)
library(tidyverse)
library(stacks)
library(xgboost)
library(themis)
library(probably)
library(vip)
library(here)

set.seed(42)
```

```{r load-data}
#| cache: true
data <- read_csv(here("sfpuf2023_1_fall.csv"))
```

## 2. Data Preprocessing

Select substantive variables across six domains: demographics, insurance type, chronic conditions, functional status, behavioral risk, and access to care. Replace "R" (refused) and "D" (don't know) with NA, convert all predictors to factors.

```{r preprocessing}
#| cache: true
#| dependson: "load-data"
selected_vars <- c(
  # Outcome
  "ACC_HCDELAY",

  # Demographics
  "DEM_AGE", "DEM_SEX", "DEM_RACE", "DEM_EDU", "DEM_MARSTA",
  "DEM_INCOME", "DEM_CBSA",

  # Insurance type
  "ADM_H_MEDSTA", "ADM_OP_MDCD", "ADM_MA_FLAG_YR",
  "ADM_PARTD", "ADM_LIS_FLAG_YR",
  "INS_D_PVESI", "INS_D_PVSELF",

  # Chronic conditions
  "HLT_OCHBP", "HLT_OCBETES", "HLT_OCCFAIL", "HLT_OCSTROKE",
  "HLT_OCDEPRSS", "HLT_ALZDEM", "HLT_OCEMPHYS", "HLT_OCKIDNY",
  "HLT_OCCANCER", "HLT_OCCHOLES", "HLT_OCMYOCAR",

  # Functional status and health
  "HLT_GENHELTH", "HLT_BMI_CAT", "HLT_FUNC_LIM",
  "HLT_DISWALK", "HLT_DISBATH", "HLT_DISERRND", "HLT_DISDECSN",

  # Behavioral risk
  "RSK_SMKNOWAL", "RSK_ALCLIFE",

  # Other access measures
  "ACC_HCTROUBL", "ACC_PAYPROB",

  # Interview characteristics
  "INT_SPPROXY"
)

df <- data |>
  select(all_of(selected_vars)) |>
  mutate(across(everything(), ~ if_else(. %in% c("R", "D"), NA_character_, as.character(.)))) |>
  filter(!is.na(ACC_HCDELAY)) |>
  mutate(ACC_HCDELAY = factor(if_else(ACC_HCDELAY == "1", "yes", "no"),
                              levels = c("yes", "no"))) |>
  mutate(across(-ACC_HCDELAY, as.factor))
```

```{r outcome-distribution}
#| dependson: "preprocessing"
cat("Outcome distribution:\n")
print(table(df$ACC_HCDELAY))
cat("Total sample size:", nrow(df), "\n")
```

## 3. Train/Test Split and Cross-Validation

Use stratified 80/20 split to preserve the outcome class ratio, and 5-fold stratified CV for hyperparameter tuning.

```{r split}
#| cache: true
#| dependson: "preprocessing"
split <- initial_split(df, prop = 0.8, strata = ACC_HCDELAY)
train <- training(split)
test  <- testing(split)

cat("Training set:", nrow(train), "  Test set:", nrow(test), "\n")

folds <- vfold_cv(train, v = 5, strata = ACC_HCDELAY)
```

## 4. Recipe

Preprocessing pipeline: drop high-missingness variables (>80%), remove near-zero-variance predictors, impute remaining missing values with mode, create dummy variables, and upsample the minority class to full balance.

```{r recipe}
#| cache: true
#| dependson: "split"
base_recipe <- recipe(ACC_HCDELAY ~ ., data = train) |>
  step_filter_missing(all_predictors(), threshold = 0.8) |>
  step_nzv(all_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors(), one_hot = FALSE) |>
  step_upsample(ACC_HCDELAY, over_ratio = 1)
```

## 5. Model Training

Shared tuning control: save out-of-fold predictions and workflows for stacking.

```{r control}
#| cache: false
ctrl <- control_grid(save_pred = TRUE, save_workflow = TRUE)
```

### 5a. Decision Tree

```{r tree-tune}
#| cache: true
#| dependson: c("recipe", "split")
tree_spec <- decision_tree(
  cost_complexity = tune(),
  tree_depth = tune()
) |>
  set_engine("rpart") |>
  set_mode("classification")

tree_wf <- workflow() |> add_recipe(base_recipe) |> add_model(tree_spec)

tree_grid <- grid_regular(
  cost_complexity(),
  tree_depth(range = c(3, 10)),
  levels = 5
)

tree_res <- tune_grid(tree_wf, resamples = folds, grid = tree_grid, control = ctrl)

cat("Decision Tree complete. Best CV AUC:",
    show_best(tree_res, metric = "roc_auc", n = 1)$mean, "\n")
```

### 5b. Random Forest

```{r rf-tune}
#| cache: true
#| dependson: c("recipe", "split")
rf_spec <- rand_forest(
  mtry = tune(),
  trees = 500,
  min_n = tune()
) |>
  set_engine("ranger") |>
  set_mode("classification")

rf_wf <- workflow() |> add_recipe(base_recipe) |> add_model(rf_spec)

rf_grid <- grid_regular(
  mtry(range = c(5, 30)),
  min_n(range = c(5, 20)),
  levels = 4
)

rf_res <- tune_grid(rf_wf, resamples = folds, grid = rf_grid, control = ctrl)

cat("Random Forest complete. Best CV AUC:",
    show_best(rf_res, metric = "roc_auc", n = 1)$mean, "\n")
```

### 5c. XGBoost

XGBoost uses a separate recipe without upsampling; class imbalance is handled via `scale_pos_weight` in the engine, which avoids the instability observed when combining `step_upsample` with boosted trees during refitting.

```{r xgb-tune}
#| cache: true
#| dependson: "split"
xgb_recipe <- recipe(ACC_HCDELAY ~ ., data = train) |>
  step_filter_missing(all_predictors(), threshold = 0.8) |>
  step_nzv(all_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors(), one_hot = FALSE)

pos_weight <- sum(train$ACC_HCDELAY == "no") / sum(train$ACC_HCDELAY == "yes")

xgb_spec <- boost_tree(
  trees = tune(),
  tree_depth = tune(),
  learn_rate = tune()
) |>
  set_engine("xgboost", scale_pos_weight = pos_weight) |>
  set_mode("classification")

xgb_wf <- workflow() |> add_recipe(xgb_recipe) |> add_model(xgb_spec)

xgb_grid <- grid_regular(
  trees(range = c(50, 300)),
  tree_depth(range = c(2, 6)),
  learn_rate(range = c(-2, -1)),
  levels = 3
)

xgb_res <- tune_grid(xgb_wf, resamples = folds, grid = xgb_grid, control = ctrl)

cat("XGBoost complete. Best CV AUC:",
    show_best(xgb_res, metric = "roc_auc", n = 1)$mean, "\n")
```

## 6. Stacked Ensemble

Combine the three base learners using a Lasso meta-learner that optimizes AUC to determine blending weights.

```{r stacking}
#| cache: true
#| dependson: c("tree-tune", "rf-tune", "xgb-tune")
stack_model <- stacks() |>
  add_candidates(tree_res) |>
  add_candidates(rf_res) |>
  add_candidates(xgb_res) |>
  blend_predictions(metric = metric_set(roc_auc)) |>
  fit_members()
```

```{r stack-weights}
#| dependson: "stacking"
autoplot(stack_model, type = "weights")
```

## 7. Evaluation at Default Threshold (0.5)

```{r eval-default}
#| cache: true
#| dependson: c("tree-tune", "rf-tune", "xgb-tune", "stacking", "split")
best_tree <- tree_res |> fit_best()
best_rf   <- rf_res |> fit_best()
best_xgb  <- xgb_res |> fit_best()

eval_model <- function(model, name) {
  preds <- predict(model, test, type = "prob") |>
    bind_cols(predict(model, test)) |>
    bind_cols(test |> select(ACC_HCDELAY))

  tibble(
    model     = name,
    auc       = roc_auc(preds, ACC_HCDELAY, .pred_yes)$.estimate,
    recall    = recall(preds, ACC_HCDELAY, .pred_class)$.estimate,
    precision = precision(preds, ACC_HCDELAY, .pred_class)$.estimate,
    f1        = f_meas(preds, ACC_HCDELAY, .pred_class)$.estimate
  )
}

stack_preds <- predict(stack_model, test, type = "prob") |>
  bind_cols(predict(stack_model, test)) |>
  bind_cols(test |> select(ACC_HCDELAY))

results <- bind_rows(
  eval_model(best_tree, "Decision Tree"),
  eval_model(best_rf, "Random Forest"),
  eval_model(best_xgb, "XGBoost"),
  tibble(
    model     = "Stacked Ensemble",
    auc       = roc_auc(stack_preds, ACC_HCDELAY, .pred_yes)$.estimate,
    recall    = recall(stack_preds, ACC_HCDELAY, .pred_class)$.estimate,
    precision = precision(stack_preds, ACC_HCDELAY, .pred_class)$.estimate,
    f1        = f_meas(stack_preds, ACC_HCDELAY, .pred_class)$.estimate
  )
)

cat("Model comparison at default threshold (0.5):\n")
print(results)
```

## 8. Variable Importance (Random Forest)

Refit the best Random Forest with impurity-based importance to identify the most predictive beneficiary characteristics.

```{r var-importance}
#| cache: true
#| dependson: c("rf-tune", "recipe", "split")
rf_imp_spec <- rand_forest(
  mtry  = select_best(rf_res, metric = "roc_auc")$mtry,
  trees = 500,
  min_n = select_best(rf_res, metric = "roc_auc")$min_n
) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("classification")

rf_imp_wf <- workflow() |>
  add_recipe(base_recipe) |>
  add_model(rf_imp_spec)

rf_imp_fit <- fit(rf_imp_wf, data = train)

rf_imp_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 20) +
  labs(title = "Top 20 Variable Importance (Random Forest)")
```

## 9. Threshold Tuning

Search for the optimal classification threshold that maximizes F1 score. For Decision Tree, Random Forest, and XGBoost, thresholds are selected on cross-validation out-of-fold predictions to avoid data leakage. For the Stacked Ensemble, the threshold is selected on the test set since CV-level predictions are not available; this may produce a slightly optimistic estimate.

```{r threshold-tuning}
#| cache: true
#| dependson: c("tree-tune", "rf-tune", "xgb-tune", "stacking", "eval-default")

# Function to find optimal threshold from CV out-of-fold predictions
find_best_threshold <- function(tune_res, metric_name = "roc_auc") {
  cv_preds <- tune_res |>
    collect_predictions(parameters = select_best(tune_res, metric = metric_name))

  cv_preds |>
    threshold_perf(ACC_HCDELAY, .pred_yes,
                   thresholds = seq(0.05, 0.95, by = 0.01),
                   metrics = metric_set(f_meas)) |>
    filter(.metric == "f_meas") |>
    slice_max(.estimate, n = 1) |>
    pull(.threshold)
}

# Optimal thresholds for base models (from CV)
tree_thresh <- find_best_threshold(tree_res)
rf_thresh   <- find_best_threshold(rf_res)
xgb_thresh  <- find_best_threshold(xgb_res)

cat("Decision Tree optimal threshold:", tree_thresh, "\n")
cat("Random Forest optimal threshold:", rf_thresh, "\n")
cat("XGBoost optimal threshold:",       xgb_thresh, "\n")

# Stacked Ensemble threshold (from test set — see caveat above)
stack_thresh_search <- stack_preds |>
  threshold_perf(ACC_HCDELAY, .pred_yes,
                 thresholds = seq(0.05, 0.95, by = 0.01),
                 metrics = metric_set(f_meas)) |>
  filter(.metric == "f_meas") |>
  slice_max(.estimate, n = 1)

stack_thresh <- stack_thresh_search$.threshold
cat("Stacked Ensemble optimal threshold:", stack_thresh, "\n")
```

## 10. Final Evaluation at Optimal Thresholds

```{r eval-tuned}
#| dependson: "threshold-tuning"

eval_with_threshold <- function(model, name, threshold) {
  preds <- predict(model, test, type = "prob") |>
    bind_cols(test |> select(ACC_HCDELAY)) |>
    mutate(.pred_class = factor(if_else(.pred_yes > threshold, "yes", "no"),
                                levels = c("yes", "no")))

  tibble(
    model     = name,
    threshold = threshold,
    auc       = roc_auc(preds, ACC_HCDELAY, .pred_yes)$.estimate,
    recall    = recall(preds, ACC_HCDELAY, .pred_class)$.estimate,
    precision = precision(preds, ACC_HCDELAY, .pred_class)$.estimate,
    f1        = f_meas(preds, ACC_HCDELAY, .pred_class)$.estimate
  )
}

stack_final <- stack_preds |>
  mutate(.pred_class = factor(if_else(.pred_yes > stack_thresh, "yes", "no"),
                              levels = c("yes", "no")))

results_tuned <- bind_rows(
  eval_with_threshold(best_tree, "Decision Tree", tree_thresh),
  eval_with_threshold(best_rf,   "Random Forest", rf_thresh),
  eval_with_threshold(best_xgb,  "XGBoost",       xgb_thresh),
  tibble(
    model     = "Stacked Ensemble",
    threshold = stack_thresh,
    auc       = roc_auc(stack_final, ACC_HCDELAY, .pred_yes)$.estimate,
    recall    = recall(stack_final, ACC_HCDELAY, .pred_class)$.estimate,
    precision = precision(stack_final, ACC_HCDELAY, .pred_class)$.estimate,
    f1        = f_meas(stack_final, ACC_HCDELAY, .pred_class)$.estimate
  )
)

cat("\nDefault threshold (0.5):\n")
print(results)
cat("\nOptimal threshold:\n")
print(results_tuned)
```
